Appendix to " New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-09-11 jso.qmd
add code as a separate file and clean up supplementary materials

To use JSO for PPGT in the `tourr` package, specify `search_f = search_jellyfish` in the guided tour. Since multiple tour paths are generated by the jellyfish, the animation function `animate_*()` will not render the tour path for JSO. To visualise path visited by each individual jellyfish, assign the animation to an object (`res <- animate_xy(...)`). This will save the bases visited by the optimiser, along with relevant metadata, as a tibble data object (see @RJ-2021-105 for more details on the data object). The bases visited for individual jellyfish can then be extracted and viewed using the `planned_tour()` function. <!-- The following code uses JSO, with 20 jellyfish and a maximum of 50 tries, on the penguins data using the `lda_pp` index.-->

```{r}
#| echo: false
#| eval: false
# clean up the penguins data
penguins_cl <- palmerpenguins::penguins |> 
  filter(!is.na(bill_length_mm)) |>
  rename(bl = bill_length_mm, bd = bill_depth_mm,
         fl = flipper_length_mm, bm = body_mass_g)
penguins_df <- penguins_cl |> select(bl:bm) |> rescale()
# run the projection pursuit with JSO 
res <- animate_xy(penguins_df, 
                  guided_tour(lda_pp(cl = penguins_cl$species),
                              search_f = search_jellyfish, 
                              n_jellies = 20, max.tries = 50))
```

@@

```{r}
#| echo: false
#| eval: false
# extract the bases from the first jellyfish
bases <- res |> filter(loop == 1) |> pull(basis) |> check_dup(0.1)
# visualise its path with a planned tour
animate_xy(data = penguins_df, tour_path = planned_tour(bases), 
           col = penguins_cl$species)
```

<!-- @fig-jso-penguins shows six frames from the tour jellyfish No.1 animation. The Gentoo species (red) was mixed with Chinstrap (yellow) initially and gradually separated out as flipper length (`fl`) gaining more weights.-->

```{r}
#| echo: false
#| eval: false
library(dplyr)
library(palmerpenguins)
library(ggplot2)
library(tourr)
check_dup <- function(bases, min_dist) {
  res <- list()
  res[[1]] <- bases[[1]]
  i <- 1; j <- 2
  while(j <= length(bases)){
    if (proj_dist(bases[[i]], bases[[j]]) >= min_dist) {
      res <- c(res, list(bases[[j]]))
      i <- i + 1; j <- j + 1
    } else{
      j <- j + 1
    }
  }
  return(res)
}

penguins_cl <- palmerpenguins::penguins |> filter(!is.na(bill_length_mm)) |>
  rename(bl = bill_length_mm, bd = bill_depth_mm,
         fl = flipper_length_mm, bm = body_mass_g)
penguins_df <- penguins_cl |> select(bl:bm) |> rescale()
res <- animate_xy(penguins_df, guided_tour(lda_pp(cl = penguins_cl$species), search_f = search_jellyfish, n_jellies = 20, max.tries = 50))
bases <- res |> filter(loop == 1) |> pull(basis) |> check_dup(0.1)
animate_xy(data = penguins_df, tour_path = planned_tour(bases), col = penguins_cl$species)
render_gif(penguins_df, planned_tour(bases), display_xy(col = penguins_cl$species), 
           here::here("figures/tour/penguins.gif"))
render( 
  penguins_df,
  tour_path = planned_tour(bases),
  dev = "png",
  display = display_xy(col = penguins_cl$species),
  frames = 100,
  here::here("figures","tour", "penguins%03d.png")
)
```

<!-- 
```{r}
#| label: fig-jso-penguins
#| fig-cap: "Visualising the tour path of a single jellyfish (No.1) in the penguins data set. The projection pursuit is optimised by the JSO with 20 jellyfishes and a maximum of 50 tries. The bases visited by jellyfish No.1 is extracted and constructed as a planned tour. The jellyfish successfully finds the structure of the data set, separating the three species of penguins."
frames <- c("001", "010", "019", "028", "036", "048")
ani <- paste0(here::here("figures/"), "tour/", "penguins", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl)
```
-->

@@

To sample bases for calculating smoothness, the required arguments include `idx` (the index function used), `data` (the dataset), `n_basis` (the number of random bases to sample), and `best` (the best basis to compute projection distance). <!--The following code samples 300 bases for using the `holes` index to find the pipe shape in the 6D `sine1000` dataset: -->

```{r}
#| echo: false
#| eval: false
library(GpGp)
library(fields)
library(tourr)
library(ferrn)
basis_smoothness <- sample_bases(
  idx = "holes", data = sine1000, n_basis = 300, 
  best = matrix(c(0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 1), nrow = 6))
basis_smoothness |> head(3)
```

The output of `sample_bases()` is a data frame with four columns: `id` (the basis ID), `basis` (the basis matrix sampled), `dist` (the projection distance to the best basis), and `index` (the index value). To speed up the computation of index values,  parallellasation is available through the `parallel = TRUE` argument.

The `calc_smoothness()` function takes the output from `sample_bases()` and fits a Gaussian process model to the index values against the projection distances to compute the smoothness metric. The starting parameter for the fit can be specified using the `start_params` argument, and additional Gaussian process arguments can be provided through `other_gp_params`. <!--The following code calculates the smoothness metric from the previously sampled bases:-->

```{r}
#| echo: false
#| eval: false
(sm_res <- calc_smoothness(basis_smoothness))
```

The smoothness metric is the smoothness parameter from the Gussian process and all details of the fit can be accessed using `attr(sm_res, "fit_res")`.

@@

The `calc_squintability()` function computes squintability using two methods: 1) parametrically, by fitting a scaled sigmoid function through non-linear least square (`method = "nls"`), and 2) non-parametrically, using kernel smoothing (`method = "ks"`). A `bin_width` argument is required to average the index values over the projection distance before the fit. <!-- The following code samples 100 bases for calculating squintability in the same scenario:-->

```{r}
#| echo: false
#| eval: false
basis_squint <- sample_bases(
   idx = "holes", data = sine1000, n_basis = 100, 
  best = matrix(c(0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 1), nrow = 6),
  step_size = 0.1, min_proj_dist = 1.5)
calc_squintability(basis_squint, method = "nls", bin_width = 0.01)
```

The output provides the estimated parameters for the scaled sigmoid function ($\theta_1$ to $\theta_4$) and the squintability metric for the parametric case. For the non-parametric case, it shows the (`max_d`), the corresponding projection distance (`max_dist`), and teh squintability metric as their products.

@@

The article is created using Quarto [@Allaire_Quarto_2022] in R [@R].
The source code for reproducing the work reported in this paper can be
found at: [https://github.com/huizezhang-sherry/paper-jso](https://github.com/huizezhang-sherry/paper-jso). Nicolas
Langrené acknowledges the partial support of the Guangdong Provincial
Key Laboratory IRADS (2022B1212010006, R0400001-22) and the UIC Start-up
Research Fund UICR0700041-22.

>>>>

To use JSO for PPGT in the `tourr` package, specify `search_f = search_jellyfish` in the guided tour. Since multiple tour paths are generated by the jellyfish, the animation function `animate_*()` will not render the tour path for JSO. To visualise path visited by each individual jellyfish, assign the animation to an object (`res <- animate_xy(...)`). This will save the bases visited by the optimiser, along with relevant metadata, as a tibble data object (see @RJ-2021-105 for more details on the data object). The bases visited for individual jellyfish can then be extracted and viewed using the `planned_tour()` function.

@@

@@

To sample bases for calculating smoothness, the required arguments include `idx` (the index function used), `data` (the dataset), `n_basis` (the number of random bases to sample), and `best` (the best basis to compute projection distance). The output of `sample_bases()` is a data frame with four columns: `id` (the basis ID), `basis` (the basis matrix sampled), `dist` (the projection distance to the best basis), and `index` (the index value). To speed up the computation of index values,  parallellasation is available through the `parallel = TRUE` argument.

The `calc_smoothness()` function takes the output from `sample_bases()` and fits a Gaussian process model to the index values against the projection distances to compute the smoothness metric. The starting parameter for the fit can be specified using the `start_params` argument, and additional Gaussian process arguments can be provided through `other_gp_params`. The smoothness metric is the smoothness parameter from the Gussian process and all details of the fit can be accessed using `attr(sm_res, "fit_res")`.

@@

The `calc_squintability()` function computes squintability using two methods: 1) parametrically, by fitting a scaled sigmoid function through non-linear least square (`method = "nls"`), and 2) non-parametrically, using kernel smoothing (`method = "ks"`). A `bin_width` argument is required to average the index values over the projection distance before the fit. The output provides the estimated parameters for the scaled sigmoid function ($\theta_1$ to $\theta_4$) and the squintability metric for the parametric case. For the non-parametric case, it shows the (`max_d`), the corresponding projection distance (`max_dist`), and teh squintability metric as their products.

@@

The article is created using Quarto [@Allaire_Quarto_2022] in R [@R].
The source code for reproducing the work reported in this paper can be
found at: <https://github.com/huizezhang-sherry/paper-jso>. The simulation data produced in Section 5 can be found at <https://figshare.com/articles/dataset/Simulated_raw_data/26039506>. Nicolas
Langrené acknowledges the partial support of the Guangdong Provincial
Key Laboratory IRADS (2022B1212010006, R0400001-22) and the UIC Start-up
Research Fund UICR0700041-22.

# Supplementary materials {-}

The supplementary materials include: 1) description of all the index functions used in the simulations and 2) the script that complements the Practicalities section for using JSO in a PPGT and calculating smoothness and squintability.

### 2024-09-11 jso.tex
add code as a separate file and clean up supplementary materials

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, colback=white, rightrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, coltitle=black, breakable, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, opacityback=0, left=2mm, bottomtitle=1mm, titlerule=0mm, arc=.35mm, toprule=.15mm, bottomrule=.15mm]

@@

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (including variance, range, smoothness, and nugget)
and scaled logistic function (\(\theta_1\) to \(\theta_4\)) for the
pipe-finding and sine-wave finding problems. The column ``smooth'' and
``squint'' represent the smoothness and squintability measures.}

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrrr>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.00 & 0.37 & 2.36 & 0.22 & 1.02 & 0.86 & 3.37 & 0.02 & 3.05\\
2 & pipe & holes & 8 & 0.00 & 0.18 & 2.19 & 0.82 & 1.01 & 0.87 & 3.26 & 0.03 & 2.96\\
3 & pipe & holes & 10 & 0.00 & 0.11 & 2.19 & 3.03 & 1.02 & 0.88 & 3.15 & 0.02 & 2.95\\
4 & pipe & holes & 12 & 0.00 & 0.15 & 2.29 & 1.58 & 1.01 & 0.88 & 3.34 & 0.00 & 3.12\\
5 & sine & MIC & 6 & 0.02 & 0.09 & 2.46 & 0.08 & 0.89 & 0.57 & 1.62 & -0.02 & 1.26\\
6 & sine & MIC & 8 & 0.02 & 0.08 & 2.64 & 0.08 & 0.93 & 0.33 & 1.31 & -0.03 & 0.77\\
7 & sine & TIC & 6 & 0.12 & 0.11 & 2.44 & 0.09 & 0.95 & 0.54 & 1.72 & -0.03 & 1.32\\
8 & sine & TIC & 8 & 0.12 & 0.10 & 2.47 & 0.09 & 0.95 & 0.56 & 1.72 & -0.03 & 1.37\\
9 & sine & dcor2d & 6 & 0.03 & 0.17 & 2.66 & 0.11 & 0.95 & 1.04 & 2.74 & -0.02 & 2.95\\
10 & sine & loess2d & 6 & 0.08 & 0.34 & 1.99 & 0.31 & 1.02 & 1.04 & 2.65 & 0.08 & 2.76\\
11 & sine & splines2d & 6 & 0.04 & 0.24 & 2.54 & 0.10 & 1.01 & 1.05 & 2.73 & -0.01 & 3.12\\
12 & sine & stringy & 6 & 0.00 & 0.01 & 1.54 & 38.17 & 1.05 & 0.01 & 254.73 & 0.05 & 2.96\\
\bottomrule
\end{tabular}
\endgroup{}

@@

To sample bases for calculating smoothness, the required arguments
include \texttt{idx} (the index function used), \texttt{data} (the
dataset), \texttt{n\_basis} (the number of random bases to sample), and
\texttt{best} (the best basis to compute projection distance).

The output of \texttt{sample\_bases()} is a data frame with four
columns: \texttt{id} (the basis ID), \texttt{basis} (the basis matrix
sampled), \texttt{dist} (the projection distance to the best basis), and
\texttt{index} (the index value). To speed up the computation of index
values, parallellasation is available through the
\texttt{parallel\ =\ TRUE} argument.

The \texttt{calc\_smoothness()} function takes the output from
\texttt{sample\_bases()} and fits a Gaussian process model to the index
values against the projection distances to compute the smoothness
metric. The starting parameter for the fit can be specified using the
\texttt{start\_params} argument, and additional Gaussian process
arguments can be provided through \texttt{other\_gp\_params}.

The smoothness metric is the smoothness parameter from the Gussian
process and all details of the fit can be accessed using
\texttt{attr(sm\_res,\ "fit\_res")}.

@@

The \texttt{calc\_squintability()} function computes squintability using
two methods: 1) parametrically, by fitting a scaled sigmoid function
through non-linear least square (\texttt{method\ =\ "nls"}), and 2)
non-parametrically, using kernel smoothing (\texttt{method\ =\ "ks"}). A
\texttt{bin\_width} argument is required to average the index values
over the projection distance before the fit.

The output provides the estimated parameters for the scaled sigmoid
function (\(\theta_1\) to \(\theta_4\)) and the squintability metric for
the parametric case. For the non-parametric case, it shows the
(\texttt{max\_d}), the corresponding projection distance
(\texttt{max\_dist}), and teh squintability metric as their products.

@@

The article is created using Quarto (Allaire et al. 2022) in R (R Core
Team 2023). The source code for reproducing the work reported in this
paper can be found at:
\url{https://github.com/huizezhang-sherry/paper-jso}. Nicolas Langrené
acknowledges the partial support of the Guangdong Provincial Key
Laboratory IRADS (2022B1212010006, R0400001-22) and the UIC Start-up
Research Fund UICR0700041-22.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\begin{tcolorbox}[enhanced jigsaw, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, opacitybacktitle=0.6, left=2mm, titlerule=0mm, toprule=.15mm, arc=.35mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, leftrule=.75mm, colframe=quarto-callout-note-color-frame, opacityback=0, breakable, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm]

@@

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (outputscale \(\eta\), lengthscale \(\ell\),
smoothness \(\nu\), and nugget \(\sigma\)) and scaled logistic function
(\(\theta_1\) to \(\theta_4\) and \(\varsigma\)) for the pipe-finding
and sine-wave finding problems. The columns \(\nu\) and \(\varsigma\)
represent the smoothness and squintability measures respectively.}

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrrr>{}r|}
\toprule
  & shape & index & d & $\eta$ & $\ell$ & $\nu$ & $\sigma$ & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & $\varsigma$\\
\midrule
1 & pipe & holes & 6 & 0.00 & 0.37 & 2.36 & 0.22 & 1.02 & 0.86 & 3.37 & 0.02 & 3.05\\
2 & pipe & holes & 8 & 0.00 & 0.18 & 2.19 & 0.82 & 1.01 & 0.87 & 3.26 & 0.03 & 2.96\\
3 & pipe & holes & 10 & 0.00 & 0.11 & 2.19 & 3.03 & 1.02 & 0.88 & 3.15 & 0.02 & 2.95\\
4 & pipe & holes & 12 & 0.00 & 0.15 & 2.29 & 1.58 & 1.01 & 0.88 & 3.34 & 0.00 & 3.12\\
5 & sine & MIC & 6 & 0.02 & 0.09 & 2.46 & 0.08 & 0.89 & 0.57 & 1.62 & -0.02 & 1.26\\
6 & sine & MIC & 8 & 0.02 & 0.08 & 2.64 & 0.08 & 0.93 & 0.33 & 1.31 & -0.03 & 0.77\\
7 & sine & TIC & 6 & 0.12 & 0.11 & 2.44 & 0.09 & 0.95 & 0.54 & 1.72 & -0.03 & 1.32\\
8 & sine & TIC & 8 & 0.12 & 0.10 & 2.47 & 0.09 & 0.95 & 0.56 & 1.72 & -0.03 & 1.37\\
9 & sine & dcor2d & 6 & 0.03 & 0.17 & 2.66 & 0.11 & 0.95 & 1.04 & 2.74 & -0.02 & 2.95\\
10 & sine & loess2d & 6 & 0.08 & 0.34 & 1.99 & 0.31 & 1.02 & 1.04 & 2.65 & 0.08 & 2.76\\
11 & sine & splines2d & 6 & 0.04 & 0.24 & 2.54 & 0.10 & 1.01 & 1.05 & 2.73 & -0.01 & 3.12\\
12 & sine & stringy & 6 & 0.00 & 0.01 & 1.54 & 38.17 & 1.05 & 0.01 & 254.73 & 0.05 & 2.96\\
\bottomrule
\end{tabular}
\endgroup{}

@@

To sample bases for calculating smoothness, the required arguments
include \texttt{idx} (the index function used), \texttt{data} (the
dataset), \texttt{n\_basis} (the number of random bases to sample), and
\texttt{best} (the best basis to compute projection distance). The
output of \texttt{sample\_bases()} is a data frame with four columns:
\texttt{id} (the basis ID), \texttt{basis} (the basis matrix sampled),
\texttt{dist} (the projection distance to the best basis), and
\texttt{index} (the index value). To speed up the computation of index
values, parallellasation is available through the
\texttt{parallel\ =\ TRUE} argument.

The \texttt{calc\_smoothness()} function takes the output from
\texttt{sample\_bases()} and fits a Gaussian process model to the index
values against the projection distances to compute the smoothness
metric. The starting parameter for the fit can be specified using the
\texttt{start\_params} argument, and additional Gaussian process
arguments can be provided through \texttt{other\_gp\_params}. The
smoothness metric is the smoothness parameter from the Gussian process
and all details of the fit can be accessed using
\texttt{attr(sm\_res,\ "fit\_res")}.

@@

The \texttt{calc\_squintability()} function computes squintability using
two methods: 1) parametrically, by fitting a scaled sigmoid function
through non-linear least square (\texttt{method\ =\ "nls"}), and 2)
non-parametrically, using kernel smoothing (\texttt{method\ =\ "ks"}). A
\texttt{bin\_width} argument is required to average the index values
over the projection distance before the fit. The output provides the
estimated parameters for the scaled sigmoid function (\(\theta_1\) to
\(\theta_4\)) and the squintability metric for the parametric case. For
the non-parametric case, it shows the (\texttt{max\_d}), the
corresponding projection distance (\texttt{max\_dist}), and teh
squintability metric as their products.

@@

The article is created using Quarto (Allaire et al. 2022) in R (R Core
Team 2023). The source code for reproducing the work reported in this
paper can be found at:
\url{https://github.com/huizezhang-sherry/paper-jso}. The simulation
data produced as described in Section 5 in can be found at
\url{https://figshare.com/articles/dataset/Simulated_raw_data/26039506}.
Nicolas Langrené acknowledges the partial support of the Guangdong
Provincial Key Laboratory IRADS (2022B1212010006, R0400001-22) and the
UIC Start-up Research Fund UICR0700041-22.

\section*{Supplementary materials}\label{supplementary-materials}
\addcontentsline{toc}{section}{Supplementary materials}

The supplementary materials include: 1) description of all the index
functions used in the simulations and 2) the script that complements the
Practicalities section for using JSO in a PPGT and calculating
smoothness and squintability.

### 2024-09-11 supplementary/index-def.qmd
add code as a separate file and clean up supplementary materials

---
title: 'Appendix to "Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit"'
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
date: last-modified
abstract: |
  Projection pursuit (PP) is a dimension reduction technique that identifies low-dimensional projections in high-dimensional data by optimising a criteria function known as the PP index. The optimisation in PP can be non-smooth and requires identifying optima with a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recent swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, this work proposes novel methods to quantify two properties of the PP index -- smoothness and squintability -- that capture the complexities inherent in PP optimisation problems. These two metrics are evaluated with JSO hyper-parameters to determine their effect on JSO success rate. The JSO algorithm has been implemented in the `tourr` package, while calculations for smoothness and squintability are available in the `ferrn` package. 
format:
  elsevier-pdf:
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
bibliography: ../bibliography.bib
---

@@

   -  column sum, row sum, and overall sum: $$a_{i \cdot} = \sum_{l = 1}^n a_{il}, \quad a_{\cdot j} = \sum_{k = 1}^n a_{kj}, \quad a_{\cdot \cdot} = \sum_{k,l = 1}^n a_{kl} \quad b_{i \cdot} = \sum_{l = 1}^n b_{il}, \quad b_{\cdot j} = \sum_{k = 1}^n b_{kj}, \quad b_{\cdot \cdot} = \sum_{k,l = 1}^n b_{kl}$$

  -   distance covariance (and variance defined similarly): $$\text{dCov}(a_{ij}, b_{ij}) = \frac{1}{n(n-3)} \sum_{i \neq j} a_{ij} b_{ij} -\frac{2}{n (n-2)(n-3)}\sum_{i = 1}^n a_{i \cdot} b_{i \cdot} + \frac{1}{n(n - 1)(n-2)(n-3)} a_{\cdot \cdot} b_{\cdot \cdot}$$

@@

>>>>

---
title: 'Appendix to " New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices"'
author: H. Sherry Zhang, Dianne Cook, Nicolas Langrené, Jessica Wai Yin Leung
date: last-modified
format: pdf
bibliography: ../bibliography.bib
---

@@

   -  column sum, row sum, and overall sum: $$a_{i \cdot} = \sum_{l = 1}^n a_{il}, \quad a_{\cdot j} = \sum_{k = 1}^n a_{kj}, \quad a_{\cdot \cdot} = \sum_{k,l = 1}^n a_{kl} $$

   $$\quad b_{i \cdot} = \sum_{l = 1}^n b_{il}, \quad b_{\cdot j} = \sum_{k = 1}^n b_{kj}, \quad b_{\cdot \cdot} = \sum_{k,l = 1}^n b_{kl}$$

  -   distance covariance (and variance defined similarly): $$\text{dCov}(a_{ij}, b_{ij}) = \frac{1}{n(n-3)} \sum_{i \neq j} a_{ij} b_{ij}  -\frac{2 \sum_{i = 1}^n a_{i \cdot} b_{i \cdot}}{n (n-2)(n-3)} + \frac{a_{\cdot \cdot} b_{\cdot \cdot}}{n(n - 1)(n-2)(n-3)}$$

@@

# Reference
